Quantizing the Bosonic String on a Loop Quantum Gravity Background

### 2024-06-27 lqg-strings.tex
iso(3,1) connection added, changed roman to arabic numerals, switched to section wise equation numbering, removed 'draft' in preamble.

\RequirePackage{snapshot}
\RequirePackage[final]{graphicx}
\documentclass[twocolumn,draft,aps,prd,10pt,nofootinbib,nobibnotes]{revtex4-2}

@@

@@

\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
% \newcommand{\H}{\mathcal{H}}

@@

As for variation \wrt the embbeding fields $X^{\lambda}$, we have
\begin{equation} \label{eqn:vary-wrt-X}
	\frac{\delta S_{I A P}}{\delta X^{\lambda}} = 0
\end{equation}
\begin{multline} \label{eqn:X-eom-inverse-area}
	F^{-} \sqrt{- g} g^{a b} \partial _{a} \partial _{b} X^{\nu} G_{\lambda \nu} + \partial _{a} \left( F^{-} \sqrt{- g} g^{a b} \right) \partial _{b} X^{\nu} G_{\lambda \nu} = \\ = \frac{1}{2} F^{-} \sqrt{- g} g^{a b} \partial _{a} X^{\mu} \partial _{b} X^{\nu} \partial _{\lambda} G_{\mu \nu} ,
\end{multline}
where
\begin{equation} \label{eqn:F-minus-def}
	F^{-} \equiv 1 - \frac{k \Delta}{2 (- g)} .
\end{equation}
Since all 2-dimensional metrics are conformally flat \cite{Schottenloher}, we use conformal gauge $g_{a b} = \e ^{\phi (x)} \eta _{a b}$ and assume flat background $G_{\mu \nu} = \eta _{\mu \nu}$, to simplify eq (\ref{eqn:X-eom-inverse-area}) into
\begin{equation} \label{eqn:X-eom-simplified}
	\eta ^{a b} \partial _{a} \partial _{b} X^{\mu} + \frac{1}{F^{-}} \eta ^{a b} \partial _{a} F^{-} \partial _{b} X^{\mu} = 0 .
\end{equation}
By expanding the definition of $F^{-}$ we get
\begin{equation} \label{eqn:coupling-X-phi}
	\eta ^{a b} \partial _{a} \partial _{b} X^{\mu} + \eta ^{a b} \frac{\left( 2 \del _{a} \phi - \del _{a} \ln (\Delta) \right)}{\dfrac{2 \e ^{2 \phi}}{k \Delta} - 1} \del _{b} X^{\mu} = 0 ,
\end{equation}
revealing a wave equation sourced by a coupling between the WS conformal factor, quantum of area field $\Delta$ and the embedding fields much akin to the dilaton field \cite{Balthazar-t-backgrounds}. We'll come back to this in \autoref{sec:emergent-dimensions}.

@@

This is where we run into a problem. There is no way to express infinitesimal translations in $d+1$ dimensions as the action of a $d+1$ dimensional matrix-valued group on a point living in $\mbb{R}^{d+1}$. In order to incorporate translations we have to resort to a trick and introduce an extra ``dummy'' dimension for the background manifold. Now we have $d+2$ embedding fields which we can write as:
\begin{equation}
	X^A := \{X^0, \vec X, X^r\} = \{X^\mu, X^r\},
\end{equation}
where $X^\mu = \{X^0, \vec X\}$ are the fields we had originally and $X^r$ is our new ``dummy'' co-ordinate. As we will see this dimension will turn out to have a very important physical interpretation in the context of Einstein-Cartan geometry.

@@

@@

However, in our case, the condition $X^r = 1$ is much stronger than the Dirichlet boundary condition, because it applies not only to the end-points of the string, but to every point of the worldsheet. In other words, the \emph{entire} string worldsheet is confined to move only along the $d+1$ dimensional submanifold $\mb{\Gamma}$. Now, one can ask what should the form of the action for a string moving in this enlarged bulk look like? Well, it will be of the same form \eqref{eqn:polyakov-gauged}, only now the worldsheet connection is an $\mf{iso}(d,1)$ valued one-form, with the embedding fields the enlarged $d+2$ dimensional set.

@@

\begin{itemize}
	\item Describe iso(d,1) connection decomposition into Lorentz piece and translation piece using \cite{Wise2010MacDowell-Mansouri} as reference.
	\item Explain why deSitter, anti-deSitter and flat space are the only consistent solutions for the enlarged bulk geometry.
	\item Recall once again the steps which led us here.
\end{itemize}

>>>>

\RequirePackage{snapshot}
\RequirePackage[final]{graphicx}
\documentclass[twocolumn,aps,prd,10pt,nofootinbib,nobibnotes]{revtex4-2}

@@

% Added Jun 27, 2024 for section wise equation numbering
\usepackage{amsmath}
\numberwithin{equation}{section}
\renewcommand{\thesection}{\arabic{section}}
%

@@

\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\ep}{\epsilon}
% \newcommand{\H}{\mathcal{H}}

@@

As for variation \wrt the embbeding fields $X^{\lambda}$, we have
\begin{equation} \label{eqn:vary-wrt-X}
	\frac{\delta S_{I A P}}{\delta X^{\lambda}} = 0
\end{equation}
\begin{multline} \label{eqn:X-eom-inverse-area}
	F^{-} \sqrt{- g} g^{a b} \partial _{a} \partial _{b} X^{\nu} G_{\lambda \nu} + \partial _{a} \left( F^{-} \sqrt{- g} g^{a b} \right) \partial _{b} X^{\nu} G_{\lambda \nu} = \\ = \frac{1}{2} F^{-} \sqrt{- g} g^{a b} \partial _{a} X^{\mu} \partial _{b} X^{\nu} \partial _{\lambda} G_{\mu \nu} ,
\end{multline}
where
\begin{equation} \label{eqn:F-minus-def}
	F^{-} \equiv 1 - \frac{k \Delta}{2 (- g)} .
\end{equation}
Since all 2-dimensional metrics are conformally flat \cite{Schottenloher2008A-Mathematical}, we use conformal gauge $g_{a b} = \e ^{\phi (x)} \eta _{a b}$ and assume flat background $G_{\mu \nu} = \eta _{\mu \nu}$, to simplify eq (\ref{eqn:X-eom-inverse-area}) into
\begin{equation} \label{eqn:X-eom-simplified}
	\eta ^{a b} \partial _{a} \partial _{b} X^{\mu} + \frac{1}{F^{-}} \eta ^{a b} \partial _{a} F^{-} \partial _{b} X^{\mu} = 0 .
\end{equation}
By expanding the definition of $F^{-}$ we get
\begin{equation} \label{eqn:coupling-X-phi}
	\eta ^{a b} \partial _{a} \partial _{b} X^{\mu} + \eta ^{a b} \frac{\left( 2 \del _{a} \phi - \del _{a} \ln (\Delta) \right)}{\dfrac{2 \e ^{2 \phi}}{k \Delta} - 1} \del _{b} X^{\mu} = 0 ,
\end{equation}
revealing a wave equation sourced by a coupling between the WS conformal factor, quantum of area field $\Delta$ and the embedding fields much akin to the dilaton field \cite{Balthazar-t-backgrounds}. We'll come back to this in \autoref{sec:emergent-dimensions}.

@@

@@

This is where we run into a problem. There is no way to express infinitesimal translations in $d+1$ dimensions as the action of a $d+1$ dimensional matrix-valued group on a point living in $\mbb{R}^{d+1}$. In order to incorporate translations we have to resort to a trick and introduce an extra ``dummy'' dimension for the background manifold. Now we have $d+2$ embedding fields which we can write as:
\begin{equation}\label{eqn:enlarged-Xs}
	X^A := \{X^0, \vec X, X^r\} = \{X^\mu, X^r\},
\end{equation}
where $X^\mu = \{X^0, \vec X\}$ are the fields we had originally and $X^r$ is our new ``dummy'' co-ordinate. As we will see this dimension will turn out to have a very important physical interpretation in the context of Einstein-Cartan geometry.

@@

However, in our case, the condition $X^r = 1$ is much stronger than the Dirichlet boundary condition, because it applies not only to the end-points of the string, but to every point of the worldsheet. In other words, the \emph{entire} string worldsheet is confined to move only along the $d+1$ dimensional submanifold $\mb{\Gamma}$. Now, one can ask what should the form of the action for a string moving in this enlarged bulk look like? Well, it will be of the same form \eqref{eqn:polyakov-gauged}, only now the worldsheet connection is an $\mf{iso}(d,1)$ valued one-form, with the embedding fields living in the enlarged $d+2$ dimensional set.

\subsection{The Einstein-Cartan String}\label{sec:einstein-cartan-string}

It is useful to see the explicit form of a $\mf{iso}(3,1)$ gauge connection. The elements of the Lie algeba of $\mf{iso}(3,1)$ have the following form \cite{Wise2010MacDowell-Mansouri}
\begin{align}\label{eqn:poincare-lie-algeba}
	T^A{}_B & := \begin{pmatrix}
		0 & b^1 & b^2 & b^3 & p^0/l \\
		b^1 & 0 & j^3 & -j^2 & p^1/l \\
		b^2 & -j^3 & 0 & j^1 & p^2/l \\
		b^3 & j^2 & -j^1 & 0 & p^3/l \\
		-\ep p^0/l & -\ep p^1/l & -\ep p^2/l & -\ep p^3/l & 0
	\end{pmatrix} \nonumber \\
	& = j^i J_i + b^i B_i + \frac{1}{l}p^a P_a,
\end{align} 
where $J_i, B_i$ are the generators of rotations and boosts respectively. $P_a := (P_0, P_i)$ are the generators of translations. The parameter $\ep$ takes values in $\{-1,0,1\}$. Strictly speaking the above expression corresponds to an $\mf{iso}(3,1)$ Lie algebra generator only when $\ep = 0$. $l$ is a length scale which must be introduced so that the components of the ``vector'' $(p^0, p^i)$ can be viewed as components of a translation vector along the $3+1$ dimensional geometry $\mb{\Gamma}$. Similarly the boosts and rotations describe
where $J_i, B_i$ are the generators of rotations and boosts respectively. $P_a := (P_0, P_i)$ are the generators of translations. The parameter $\ep$ takes values in $\{-1,0,1\}$.

Strictly speaking the above expression corresponds to an $\mf{iso}(3,1)$ Lie algebra generator only when $\ep = 0$. $l$ is a length scale which must be introduced so that the components $(p^0, p^i)$ can be viewed as components of a ``translation vector'' living in the tangent space of the $3+1$ dimensional geometry $\mb{\Gamma}$. Similarly the boosts and rotations generators are for objects living in $\mb{\Gamma}$.

Let us write down the gauged Polyakov action for the Poincaré lie algebra valued connection $\mc{A}_a$ whose components are given in \eqref{eqn:poincare-lie-algeba}:
\begin{equation}\label{eqn:poincare-polyakov}
	S'_{P} = -\frac{T}{2} \int \diff^2 x \sqrt{-g} g^{ab} \mc{D}_a X^A \mc{D}_b X^B \eta_{AB},
\end{equation}
where now the $X^A = \{X^\mu, X^r\}$ (as mentioned previously in \eqref{eqn:enlarged-Xs}) and $\eta_{AB}$ is the flat metric on the $d+2$ enlarged bulk manifold $\mb{M}$ and the covariant derivative is of the form defined in \eqref{eqn:covariant-deriv-v1}, but only now acting on $X^A$ instead of on $X^\mu$:
\begin{equation}\label{eqn:covariant-deriv-v2}
	\mc{D}_a X^A(\tau , \sigma) = \partial_a X^A + k \mc{A}_{a}^A{}_B X^B.
\end{equation}
Now, we will expand out the action \eqref{eqn:poincare-polyakov} and express it in terms of the $X^\mu$ and $X^r$ separately. It is clear that the resulting expression will be sum of three terms:
\begin{align}
	S'_{P} & = -\frac{T}{2} \int \diff^2 x \sqrt{-g} g^{ab} \left( \mc{D}_a X^\mu \mc{D}_b X^\nu \eta_{\mu\nu} + \right. \nonumber \\
	& \left. \mc{D}_a X^\mu \mc{D}_b X^r \eta_{\mu r} + \mc{D}_a X^r \mc{D}_b X^r \eta_{rr} \right)
\end{align}

@@

\begin{itemize}
	\item Explain why deSitter, anti-deSitter and flat space are the only consistent solutions for the enlarged bulk geometry.
	\item Recall once again the steps which led us here.
\end{itemize}
